Typed Clojure: A Practical Optional Type System for Clojure

### 2012-09-11 dissertation/future-work.tex
Rearrange into a file per chapter

\documentclass{cshonours}
\bibliographystyle{acm}

%\usepackage{biblatex}

\title{Typed Clojure}
\author{Ambrose Bonnaire-Sergeant}

\keywords{Type systems}

\bibliography{bibliography}

\usepackage{color}
\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\begin{document}
\maketitle

\chapter{Introduction}

This dissertation describes Typed Clojure, an optional static type system for Clojure.

% Based on Typed Racket

\chapter{Typed Clojure through Examples}

% Hello world
This section introduces Typed Clojure with example code. An attempt is
made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is assumed.

We begin with the obligatory \emph{Hello world} example.

\begin{lstlisting}[caption=Hello world]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]]))

(println "Hello world")
\end{lstlisting}

The \lstinline|clojure.core/ns|
\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
invocation declares the current namespace to be \lstinline|typed.test.hello-world|
and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
\lstinline|typed.core/check-ns| to be referenced locally. Other than this
dependency, this is identical to the untyped \emph{Hello world}.
% footnote: check-ns used to initiate type checking usually at REPL

More complex code require extra annotations to type check:

\begin{lstlisting}
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Sam's dissertation TODO}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}
\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.

\section{Datatypes and Protocols}

We can annotate datatype and protocol definitions similarly.

\begin{lstlisting}
(ns typed.test.deftype
  (:require [typed.core 
             :refer [check-ns ann-datatype
                     tc-ignore ann-protocol AnyInteger]]))

(ann-protocol Age 
  :methods
  {age [Age -> AnyInteger]})
(tc-ignore
  (defprotocol Age
    (age [this])))

(ann-datatype Person 
  [[name :- String]
   [age :- AnyInteger]])
(deftype Person [name age]
  Age
  (age [this] age))

(age (Person. "Lucy" 34))
\end{lstlisting}

\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
is for this function.

Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.

\lstinline|clojure.core/deftype|
defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
in the current namespace with a number of fields and methods. 
\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
In this example, we create a datatype \lstinline|typed.test.person.Person|
with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
method from protocol \lstinline|typed.test.person/Age|.

Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
in the order they are passed to \lstinline|deftype|
\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
instance, setting the fields to their corresponding positional arguments.
Typed Clojure checks the datatype constructor to be the equivalent of 
\lstinline|[String AnyInteger -> Person]|.

Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.

\section{Polymorphism}

Typed Clojure supports F-bounded polymorphism. All type variables may have
upper and lower bounds.

Typed Clojure parameterises some of Clojure's data structures. For example,
the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
because Integer is a subtype of Number.}.

\begin{lstlisting}
...
(ann to-set 
     (All [x]
       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
(defn to-set [a]
  (set a))
...
\end{lstlisting}

In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
\lstinline|All| introduces a set of type variables to the body of a type,
here \lstinline|x| is used to define a relationship between the input type and return type.

\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
this as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.

\section{Heterogeneous Maps}

Where particularly object-oriented languages reach for objects, Clojure
utilises maps. Clojure provides a map literal using curly braces. For example,
\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included for readability.

Typed Clojure provides a heterogeneous map type, restricted to 
maps with singleton Keyword keys. This restriction is reflected
in the syntax for defining heterogeneous map types.

\begin{lstlisting}
...
(ann config
     (HMap {:file String,
            :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

This example checks \lstinline|config| to be a heterogeneous map
with \lstinline|:file| and \lstinline|:ns| keys, with values of
type \lstinline|String| and \lstinline|Symbol| respectively.

Heterogeneous vector and seq types are also provided.

\section{Multi-Arity Functions}

All Clojure functions support multiple arities, allowing different
behaviour for different number of arguments.

\begin{lstlisting}
(ns typed.test.poly
  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
  (:import [clojure.lang Seqable]))

(ann repeatedly'
     (All [x]
       (Fn [[-> x] -> (Seqable x)]
           [AnyInteger [-> x] -> (Seqable x)])))
(defn repeatedly'
  "Takes a function of no args, presumably with side effects, and
  returns an infinite (or length n if supplied) lazy sequence of calls
  to it"
  ([f] (lazy-seq (cons (f) (repeatedly' f))))
  ([n f] (take n (repeatedly' f))))
\end{lstlisting}

This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
The type variable \lstinline|x| links the return type of the generator function to
the resulting \lstinline|Seqable| instance.

When checking the definition of multi-arity functions, Typed Clojure
finds the matching function types for a particular arity by number of arguments
and requires each relevant type to check successfully.

\section{Variable-Arity Functions}

Clojure functions can take an optional \emph{rest} argument.

\subsection{Uniform Variable-Arity}

\subsection{Non-uniform Variable-Arity}

\section{Occurrence Typing}

Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
Occurrence typing is used to refine the known types for immutable local bindings 
down conditional branches.

\begin{lstlisting}
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

To check this example, occurrence typing infers type information based on the result of each test.
In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
to \lstinline|nil| from \lstinline|(U nil Number)|.

\section{Java Interoperability}

Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
\footnote{Compatibility with Generic Java is planned future work}.

The key to safe Java interop is the treatment of Java's \emph{null}.
\emph{null} is a subtype to all reference types
represented in Clojure by \lstinline|nil|. Unlike Java's type system
Clojure explicitly separates \emph{null} and reference types, allowing
Typed Clojure to express
\emph{nullable}
\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
positions.

\begin{lstlisting}
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

(ann f File)
(def f (File. "a"))

(ann prt (U nil String))
(def prt (.getParent ^File f))

(non-nil-return java.io.File/getName :all)
(ann nme String)
(def nme (.getName ^File f))

\end{lstlisting}

This example shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.

Typed Clojure checks calls to Java constructors by requiring the provided
arguments be acceptable input to at least one constructor for that Class.
In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.

Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
The \emph{java.io.File} instance method \emph{getParent}
is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
a valid approximation of the method as \emph{getParent} returns \emph{null} 
``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
On the other hand, the instance method \emph{getName} always returns an
instance of \emph{java.lang.String}, so we set the return position of
\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
to override all arities of that method.}.

\chapter{Design Goals}

\section{Safer Java Interop}
\section{Optional Type Checking}
\section{Explicit Typing}
\section{No new idioms}

\section{Local Type Inference}

Similar to Typed Racket, type inference is based on Local Type Inference
by Pierce and Turner.

\section{F-bounded Polymorphism}

Typed Clojure includes support for F-bounds on type variables, as an extension
to Local Type Inference.

An extra environment from type variables to bounds is kept until after inference,
after which each unground bound is substituted with the inferred types and the
subtyping relationships are checked.

\section{Heterogeneous Maps}
% keyword keys for now, generally encouraged, fast
% Some heterogeneous maps have string args, eg. from web frameworks

\section{Java Interoperability}
% Args non-nil, return nilable
% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)

\subsection{Primitive Arrays}
% Primitive arrays are covariant, which is well-known to be statically unsound.
% We cannot trust the type signature of any Java method or field.
% Separate read and write types into two parameters.

\subsection{Interaction with null}
% null is explicitly expressible in my type system
% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
%

\subsection{Generic Java}
% Existential types are needed to fully express Generic Java types in Typed Clojure
% F-bounded polymorphism supported

\section{Occurrence Typing}
% interesting paths
% - count path element
% - class path element
% - keyword path element

\section{Variable-arity Polymorphism}
% Problematic areas
% - flattened paired arguments. assoc, hash-maps, array-map
% - complex argument dependencies. comp, partial

\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}

Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
representing a sequence of types, allowing non-uniform variable parameters.
This enables \lstinline|map| and other function with complex arguments to be typed.
Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
Typed Racket, and is a key feature of Typed Racket's implementation.

Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
associates key-value pairs to a target map. It takes one or more key-value pairs and
applies them left-to-right to the target map. The key problem from a typing perspective
is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
higher-order function.
\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
dotted parameters represent an unrestricted number of types, where we require an even number.

The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
Firstly, subtyping between different kinds of dotted and rest
\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
difficult to express.
Thirdly, higher-order functions taking functions with dotted arguments commonly 
The first issue requires careful design and implementation.
The second and third points require new constructs, in the form of \emph{projections}
and abstractions over \emph{dotted} parameters, respectively.

\section{Implementation}

\subsection{Composite Propositions}

\subsection{Non-hygienic Macro Expansion and Filters}
% need to remove bindings as they fall out of scope

\section{Future Work}

\subsection{Multimethods}
% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

\subsection{Records}
% A composition of existing features: Datatypes and Heterogeneous Maps

@@

%
%\begin{mathpar}
%\inferrule*[right=B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\inferrule*[right=B-Assoc]
%  {\rho \vdash e_m \downarrow mi \\
%   \rho \vdash e_k \downarrow k \\
%   \rho \vdash e_v \downarrow v \\
%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}

%\infrule[B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\infrule[B-Assoc]

%\end{mathpar}

\end{document}

>>>>

\chapter{Future Work}

\section{Multimethods}
% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

\section{Records}
% A composition of existing features: Datatypes and Heterogeneous Maps

@@

### 2012-09-11 dissertation/intro.tex
Rearrange into a file per chapter

\documentclass{cshonours}
\bibliographystyle{acm}

%\usepackage{biblatex}

\title{Typed Clojure}
\author{Ambrose Bonnaire-Sergeant}

\keywords{Type systems}

\bibliography{bibliography}

\usepackage{color}
\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\begin{document}
\maketitle

@@

This dissertation describes Typed Clojure, an optional static type system for Clojure.

% Based on Typed Racket

\chapter{Typed Clojure through Examples}

@@

\section{Datatypes and Protocols}

@@

\section{Polymorphism}

@@

\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
this as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.

\section{Heterogeneous Maps}

@@

\section{Multi-Arity Functions}

All Clojure functions support multiple arities, allowing different
behaviour for different number of arguments.

@@

\section{Variable-Arity Functions}

@@

\subsection{Uniform Variable-Arity}

\subsection{Non-uniform Variable-Arity}

\section{Occurrence Typing}

Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
Occurrence typing is used to refine the known types for immutable local bindings 
down conditional branches.

@@

\section{Java Interoperability}

@@

Typed Clojure checks calls to Java constructors by requiring the provided
arguments be acceptable input to at least one constructor for that Class.
In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.

@@

\chapter{Design Goals}

\section{Safer Java Interop}
\section{Optional Type Checking}
\section{Explicit Typing}
\section{No new idioms}

\section{Local Type Inference}

Similar to Typed Racket, type inference is based on Local Type Inference
by Pierce and Turner.

\section{F-bounded Polymorphism}

Typed Clojure includes support for F-bounds on type variables, as an extension
to Local Type Inference.

An extra environment from type variables to bounds is kept until after inference,
after which each unground bound is substituted with the inferred types and the
subtyping relationships are checked.

\section{Heterogeneous Maps}
% keyword keys for now, generally encouraged, fast
% Some heterogeneous maps have string args, eg. from web frameworks

\section{Java Interoperability}
% Args non-nil, return nilable
% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)

\subsection{Primitive Arrays}
% Primitive arrays are covariant, which is well-known to be statically unsound.
% We cannot trust the type signature of any Java method or field.
% Separate read and write types into two parameters.

\subsection{Interaction with null}
% null is explicitly expressible in my type system
% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
%

\subsection{Generic Java}
% Existential types are needed to fully express Generic Java types in Typed Clojure
% F-bounded polymorphism supported

\section{Occurrence Typing}
% interesting paths
% - count path element
% - class path element
% - keyword path element

\section{Variable-arity Polymorphism}
% Problematic areas
% - flattened paired arguments. assoc, hash-maps, array-map
% - complex argument dependencies. comp, partial

\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}

Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
representing a sequence of types, allowing non-uniform variable parameters.
This enables \lstinline|map| and other function with complex arguments to be typed.
Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
Typed Racket, and is a key feature of Typed Racket's implementation.

Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
associates key-value pairs to a target map. It takes one or more key-value pairs and
applies them left-to-right to the target map. The key problem from a typing perspective
is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
higher-order function.
\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
dotted parameters represent an unrestricted number of types, where we require an even number.

The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
Firstly, subtyping between different kinds of dotted and rest
\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
difficult to express.
Thirdly, higher-order functions taking functions with dotted arguments commonly 
The first issue requires careful design and implementation.
The second and third points require new constructs, in the form of \emph{projections}
and abstractions over \emph{dotted} parameters, respectively.

\section{Implementation}

\subsection{Composite Propositions}

\subsection{Non-hygienic Macro Expansion and Filters}
% need to remove bindings as they fall out of scope

\section{Future Work}

\subsection{Multimethods}
% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

\subsection{Records}
% A composition of existing features: Datatypes and Heterogeneous Maps

\subsection{Other Clojure Dialects}
% unify ASTs for easier porting

\subsection{Linter}

%
%\begin{mathpar}
%\inferrule*[right=B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\inferrule*[right=B-Assoc]
%  {\rho \vdash e_m \downarrow mi \\
%   \rho \vdash e_k \downarrow k \\
%   \rho \vdash e_v \downarrow v \\
%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}

%\infrule[B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\infrule[B-Assoc]

%\end{mathpar}

\end{document}

>>>>

@@

\section{Thesis}

\emph{It is practical and useful to design and implement an optional typing system 
for the Clojure programming language that allows Clojure programmers to continue 
using idioms and style found in current Clojure code.}

\section{Motivation}

% Type Systems for lisp?
% Why type systems?
% Why Clojure?

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. This idea not new,
but recent attempts are noteworthy for their successful use of bidirectional checking.
Instead of inferring types the algorithm relies on programmer annotations.

Clojure is a dynamically typed language running on the JVM. It emphasises
functional programming with immutable data structures
and provides direct interoperability with JVM languages.

Recently languages have been created to or been modified to support aspects
of static and dynamic typing.
Dart is fundamentally dynamically typed but offers a form of optional
static typing. Haskell recently supports the ability to delay static type errors 
to runtime, removing the need for the program to type check before running.
Scala can emulate some features of dynamic languages with Dynamic Scala.

\subsection{What kind of type system does Typed Clojure provide?}

% What are types? No real agreement .. but here is my definition..
% Every programmer thinks they know what they are. More than one concept
% in the literature.
% Static vs. dynamic
% intrinsic vs. extrinsic

% is it even a type system?
% type refinements vs ordinary types
% extrinsic vs intrinsic type systems

% refinements check invariants of programs
% while ordinary static types check whether programs
% are basically meaningful - Rowan + Frank P
% Page 1 Rowan's PhD
An ordinary static type system is used to check whether programs are basically
meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
static type system relies on programs being well-typed for runtime semantics to be defined.
This means such programs must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
are not defined by the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a static type system that supports exact one type.

A type checker ensures invariants hold complains if code
does not conform to its documented type. Powerful type systems like
Haskell's and Scala's are useful when reasoning about higher-order code.

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. Runtime contracts are a popular alternative.
Clojure takes this approach, providing easy syntax for defining pre-
and post-conditions \footnote{Inspired by Eiffel}.

% clojure is dynamically typed
% currently few tools to statically reason about complex higher-order code
% conduit based on streams, uses continuations
% core.logic based on monads
% Static types are useful for reasoning for such things
% Strong statically typed languages like Haskell use monads regularly

% strength of Clojure is concision
% typically, concise static languages are built concurrently with their type system
% Eg. Haskell, ML flavoured languages
% Counterpoint: Typed Racket preserves racket idioms, adds static typing

\section{Typed Clojure through Examples}

@@

\subsection{Datatypes and Protocols}

@@

\subsection{Polymorphism}

@@

\lstinline|(U nil (Seqable x))| is a common type in Typed Clojure, read as the union
of the type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
it as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.

\subsection{Heterogeneous Maps}

@@

\subsection{Multi-Arity Functions}

Clojure function objects support multiple arities, allowing dispatch
on the number of supplied arguments.

@@

\subsection{Variable-Arity Functions}

@@

\subsubsection{Uniform Variable-Arity}

\subsubsection{Non-uniform Variable-Arity}

\subsection{Occurrence Typing}

Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
Occurrence typing is used to refine types for immutable local bindings 
as programs follow conditional branches.

@@

\subsection{Java Interoperability}

@@

Typed Clojure checks calls to Java constructors by requiring the provided
arguments be acceptable input to at least one constructor for that Class.
In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
argument, so \lstinline|(File. "a")| is type safe. Java constructors never
return \emph{null}, so Typed Clojure assigns the return type to be \lstinline|File|.
This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.

@@

### 2012-09-11 dissertation/lit-review.tex
Rearrange into a file per chapter

\documentclass[12pt, a4paper]{article}
\usepackage{biblatex}
\usepackage{fancyvrb}

\usepackage{color}
\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  %showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=b,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\bibliography{bibliography}
\title{A Practical Optional Type System for Clojure, Literature Review}
\author{Ambrose Bonnaire-Sergeant (20350292)}

\begin{document}
\maketitle

\begin{abstract}
Clojure is a practical, dynamically typed language running on the Java Virtual Machine.
I intend to design and implement a prototype optional static type 
system for Clojure, with practical use being a priority. Devising a static type system 
requires a combination of components that each adds some utility
without significantly complicating the overall implementation.
I look at several novel inventions created as a result 
of the Typed Racket project including
non-uniform variable-arity polymorphism and occurrence typing.
Typed Racket and Scala both use flavours of local type inference
which I compare and judge for applicability to this project.
Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
\end{abstract}

\tableofcontents

@@

\printbibliography

\end{document}

>>>>

% TODO
% What does this Lit review cover?
% Why things are included/not included?
% Background
% Soft typing
% See Sam's dissertation
% Lisp type system background (197x)
% Why was it so hard?
% - bidirectional checking vs. global inference

\chapter{Literature Review}

%\begin{abstract}
%Clojure is a practical, dynamically typed language running on the Java Virtual Machine.
%I intend to design and implement a prototype optional static type 
%system for Clojure, with practical use being a priority. Devising a static type system 
%requires a combination of components that each adds some utility
%without significantly complicating the overall implementation.
%I look at several novel inventions created as a result 
%of the Typed Racket project including
%non-uniform variable-arity polymorphism and occurrence typing.
%Typed Racket and Scala both use flavours of local type inference
%which I compare and judge for applicability to this project.
%Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
%the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
%\end{abstract}

@@
